---
title: "Diamonds distribution dashboard"
output: flexdashboard::flex_dashboard
runtime: shiny
---

```{r setup, include = FALSE}
library(flexdashboard) 
library(tidyverse)
knitr::opts_chunk$set(fig.width = 5)
```

# Statistics

## Column 1 {.tabset}

### Cut

How many diamonds of each cut quality do we have in the data set?

```{r}
source("Main_Wrangle.R")
ggplot(data=team_stats_season) +
  geom_point(mapping = aes(x=year, y=average_points, color = Team))+
  theme(legend.position = "none")
plotly::ggplotly()
```


## Custom Statistics


```{r, echo=FALSE}
# Create a Shiny app
library(shiny)
library(DT)

# Read the data
source("Main_Wrangle.R")

# Define the UI for the Shiny app
ui <- fluidPage(
  titlePanel("Custom Statistics"),
  
  # Year selection
  selectInput("year", "Select Year", choices = unique(team_stats_season$year)),
  
  # Team selection
  selectInput("team", "Select Team", choices = unique(team_stats_season$Team)),
  
  # Display statistics
  dataTableOutput("stats_table")
)

# Define the server for the Shiny app
server <- function(input, output) {
  
  # Filter data based on user selections
  filtered_data <- reactive({
    team_stats_season %>%
      filter(year == input$year, Team == input$team)
  })
  
  # Create a table for statistics
  output$stats_table <- renderDataTable({
    stats <- filtered_data() %>%
      select(win_percentage, average_points)
    datatable(stats)
  })
}

# Run the Shiny app
shinyApp(ui, server)
