---
title: "Exploration of Euroleague basketball 2007-2020"
author: "Deli Alan, Rashiti Lirim and Reis Matos Dylan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: pdf
editor: visual
output: html_document
---

# Introduction

We live in a world where the use of statistics and data analysis is of paramount importance to gain a comprehensive understanding of various subjects, allowing individuals to grasp the broader picture of what is happening around them. In the realm of sports, this emphasis on statistics is no exception. In contemporary sports, statistics play a pivotal role in ranking players and determining their relative skills and performance. Furthermore, teams increasingly rely on statistics to gain insights into the strategies employed by opposing teams, enabling them to develop effective counterstrategies for a higher chance of victory.

Beyond the competitive aspect, statistics and data analysis are crucial in sports to track how the game evolves over time. This data-driven approach aids in adapting to changing trends and styles in the sport, as well as in refining player training and development to align with the modern game.

Given our collective passion for sports, our group sought to identify a subject that would resonate with each member. We stumbled upon a remarkable database containing detailed information about basketball games in the EuroLeague, which piqued our interest (further details provided in the Data section of the report).

Recognizing the global significance of the NBA (National Basketball Association) in the world of basketball, as well as its substantial influence on other basketball organizations, we decided to focus our project on analyzing how the playing style has evolved over the years, with a particular emphasis on three-point shooting. We aimed to explore how changes in playing style impact game outcomes and sought to investigate various other statistical trends. Moreover, we intended to shed light on how the game's dynamics shift as it progresses, with strategies adapting based on the game's context---whether it's a closely contested match or not. This analysis also delves into foul statistics in different game scenarios. Lastly, we aimed to provide a comprehensive overview of individual player statistics and their impact on their respective teams.

```{r}
#| echo: false
#| include: false 
source("Main_Wrangle.R")

```

# Data

We selected our dataset from Kaggle, a prominent data science competition platform renowned for hosting a wide array of datasets.

The dataset in question provides comprehensive play-by-play information for Euroleague basketball games, spanning from 2007 through 2020. Each entry within this dataset encapsulates a specific game event, encompassing details such as the game minute, involved player and his respective team, action type, and whether it entails a two or three-point shot attempt, along with its successful outcome or miss. This invaluable dataset serves as a foundational resource for analyzing and understanding the intricacies of Euroleague basketball across the specified time period. This extensive dataset, containing over 1.6 million entries, required thorough wrangling to extract and uncover interesting stories.

As an example, here is the first entry of this dataset.

```{r}
#| echo: false 
head(euroleague, 1)
```

Given the wealth of data at our disposal, the process of discerning relevance and assembling pertinent information was essential. We meticulously organized the voluminous play-by-play entries, systematically aggregating them into coherent units of games. Subsequently, these games were further categorized into distinct years or seasons, thereby facilitating the extraction of comprehensive statistical insights related to teams and players. This strategic approach not only streamlines the analytical process but also enables a more granular exploration of performance metrics across the evolving landscape of Euroleague basketball.

Parler de media white et media black team. 2007 + 2015-2020 pas de dunk lay-up. Enlever les NA Renommer les équipes, noms presque similaires Créer des nouvelles colonnes (total point, winner, etc.)

# Exploratory data analysis

blabla

# Analysis

It is a common belief that having one of the best scorers in a sports league can significantly improve a team's chances of winning games throughout the season. To investigate the statistical validity of this belief, we conducted an analysis where we aggregated data for each player, calculating the total number of points they scored over the course of a season and then dividing it by the number of games they played. This allowed us to obtain the average number of points a player scored per game.

To further explore the relationship between a player's performance and their team's success, we employed the ***left_join*** function to merge the player's dataset with the team's dataset. This merging process enabled us to determine the win percentage of the top scorer for a given season and examine whether there is a correlation between a player's individual performance and their team's overall success.

By conducting this analysis, we aim to gain insights into the impact of top scorers on team performance and potentially uncover statistical evidence to support or refute the commonly held notion that a top scorer can significantly contribute to a team's success in a sports league.

```{r}
#| echo: false
#| message: false
#| hide: true

best_in_class_player <- stat_per_game_player |>
  group_by(year)|>
  filter(avg_points == max(avg_points, na.rm = TRUE))|>
  ungroup()

ggplot(stat_per_game_player) +
  geom_point(mapping = aes(x = year, y = avg_points, color = PLAYER)) +
  geom_smooth(mapping = aes(x = year, y = avg_points)) +
  labs(title = "Average number of points scored per game",
        subtitle = str_wrap("The points represent different players"),
        x = "Season year",
        y = "Number of points scored") +
  theme_minimal() +
  theme(legend.position = "none")

ggplot(best_in_class_player, aes(x = year, y = win_percentage)) +
  geom_line() +
  geom_point(mapping = aes(x= year, y = win_percentage)) +
  labs(title = "Win percentage of the best statistic player's team",
        x = "Season year",
        y = "Win percentage") +
  ggrepel::geom_label_repel(aes(x = year, y = win_percentage, label = Team),
                            data = best_in_class_player,
                            box.padding = 0.4,
                            size = 1.5) +
  theme_minimal() +
  theme(legend.position = "none")
```

These figures provide insights into the distribution of the average number of points scored by players across different seasons. A notable observation is the consistent trend, which is suspended around an average of approximately five points per game over the years. Despite the presence of players who excel with significantly higher points per game, the density of players clustered around the five-point mark seems to prevent a substantial overall improvement in the average points scored by individual players.

In the second graph, we examine the winning percentage of the team with the top scorer (the highest points in the first plot). Surprisingly, the data reveals that having the top scorer on one's team during a particular year does not emerge as a significant variable for predicting victories. This suggests that the presence of a top scorer alone may not be a strong determinant of a team's success, and other factors likely play a more substantial role in achieving wins.

We also examined the data regarding the number of defensive and offensive rebounds made by players in each season. Our interpretation of this data aligns with our earlier observations regarding the average number of points scored by players. It is evident that achieving success in basketball encompasses a multitude of factors, including tactics, psychology, and physical fitness. However, our analysis underscores that individual player performance, in terms of rebounds or points scored, may not, on its own, carry enough statistical significance to reliably predict a team's chances of winning. The path to victory in basketball appears to be more complex and multifaceted, involving a combination of various factors beyond individual player statistics.



# Conclusion

blabla
